Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the \texttt{eechidna} \texttt{R} package.

### 2019-03-31 draft-paper.tex
pdf and tex files

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the eechidna R package.},
            pdfkeywords={blah, blah},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\begin{document}
\maketitle
\begin{abstract}
A brief summary of our ideas
\end{abstract}
\begin{keywords}
blah, blah
\end{keywords}

@@

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3\% and 53.5\% (in favour of the Liberal party) over this period. This study explores how electoral characteristics relate to two-party preference, and whether their effects have changed over time. Electoral socio-demographics are derived from the Census, and vote counts are obtained from federal elections.

Joining these two data sources is problematic as there in an inherent asynchronicity in the two events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of obtaining appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see \textcite{DavisStimson98}, \textcite{Stimson06}, \textcite{Liao09} and \textcite{Stimson09}). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are imputed using a spatio-temporal imputation that combines areal interpolation \autocite{Goodchild1993} and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in section \ref{data}. All data and associated documentation relating to this procedure are available in the \texttt{eechidna} \texttt{R} package \autocite{eechidna}, providing a resource for future analysis .

Previous work on modelling Australian federal elections have found that aggregate socio-demographics are relatively good predictors of voting outcomes. \textcite{Forrest01} does this using multiple regression of the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998. \textcite{Stimson06}, \textcite{Stimson09} and \textcite{Stimson12} use principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. On the contrary, \textcite{Stimson09} models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of swing vote can be explained by Census data. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed, where each federal elections is modelled with a cross-sectional data set. The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from those previous.

The cross-sectional data set for each election consists of the two-party preferred vote (response variable), and socio-demographic variables (explanatory variables) that characterise each electorate. To obtain these cross-sections, socio-demographic variables are first standardized, and then principal components are used to group variables into ``factors''. To account for the inherent spatial structure of the data, a spatial error model is fit for each election. These steps are discussed in section \ref{modelling}. In section \ref{results} inference is conducted on the models to see which effects are significant, how effects change over time and which electorates have abnormal voting behaviour.

@@

The voting outcome of interest is the electoral two-party preferred vote, which is provided by the Australian Electoral Commission (AEC) for the 2001, 2004, 2007, 2010, 2013 and 2016 elections via the AEC Tally Room. The AEC divide Australia into 150 regions called electorates, with each corresponding to a single seat in the House of Representatives. Voting is compulsory in Australia, and each voter assigns a numbered preference to each available candidate in their electorate. The two-party preferred vote is determined by a tally of these preferences where, by convention, only the ranks of the Labor and Liberal candidates are considered. This is recorded as a percentage preference in favour of the Liberal party.

Socio-demographic variables are derived from the Census of Population and Housing (Census), which is a survey of every household in Australia, recording information such as age, gender, ethnicity, education level and income. There have been four Censuses in the 21st century, being that in 2001, 2006, 2011 and 2016. The Australian Bureau of Statistics (ABS) conducts the Census and publishes aggregated information. The ABS approximation of electorates at the time of the Census is chosen. From this aggregate information, 67 socio-demographic variables are computed for each of the electorates.

Raw data is sourced online from the AEC and ABS websites in .csv and .xlsx files. The format of these files change over the years, making extracting the appropriate information a big task. The functions available in the \texttt{dplyr} \autocite{dplyr} and \texttt{readxl} \autocite{readxl} \texttt{R} packages are very useful, as they provide fast consistent tools for data manipulation and functions to import .xlsx files (respectively). The 2001 and 2006 Census data are however published in a format where each electorate has a separate document, making it difficult to use the \texttt{dplyr} tools and instead cells have to be selected from each individual file to construct the desired variables. All scripts required for the data wrangling process can be found in the github repository for the \texttt{eechidna} \texttt{R} package \autocite{eechidna}, along with the raw data. The \texttt{eechidna} package makes this study entirely reproducible and provides a resource to help wrangle data for future Censuses and elections, when they become available.

@@

Between 2001 and 2016 there were six elections and four Censuses (see Figure \ref{fig:timeline}). Electoral boundaries are redistributed regularly by the AEC, meaning that only in the years where both a Census and election occur will the boundaries match - the case for the 2001 and 2016 election. Therefore, for the four elections between 2004 and 2013, both temporal and spatial differences in electorates need to be accounted for when joining the electoral two-party preferred vote with Census data. For these elections a spatio-temporal imputation method is employed to obtain electoral socio-demographics. This method uses Census information from both before and after the election of interest.

@@

{\centering \includegraphics[width=0.9\linewidth]{skeleton_files/figure-latex/timeline-1}

@@

To account for spatial differences, the piece-wise approximation method in \textcite{Goodchild1993} is adopted. Consider a map of source zones \(s = 1,...,S\), for which socio-demographic information is available, and a set of target zones \(t = 1,...,T\) for which information is to be imputed. In this context the map of electoral boundaries at the time of a Census would be the source zones, and the boundaries at the time of the election would be the target zones. Denote the area of intersection between source zone \(s\) and target zone \(t\) as \(A_{s,t}\), the population of the source zone \(s\) as \(U_s\), and the population of intersection between source zone \(s\) and target zone \(t\) as \(P_{s,t}\).

Compute each \(A_{s,t}\) and estimate population of the intersection:

\[\hat{P}_{s,t} = \frac{U_s*A_{s,t}}{\sum_{t=1}^T A_{s,t}}\]
This assumes that populations are uniformly distributed within each source zone.

In order to calculate socio-demographic information for each of the target zones, a weighted average is taken using the estimated population as weights. Denote a given Census variable for the target zone \(C_t\), and the same Census variable for the source zone \(D_s\):

\[\hat{C}_t = \frac{\sum_{s=1}^{S}{D_s*\hat{P}_{s,t}}}{\sum_{s=1}^{S}{\hat{P}_{s,t}}}\]
This assumes that each individual in a source zone assumes the aggregate characteristics of the zone.

Applying this to each of the target zones addresses the spatial component, as it imputes the required socio-demographic for the desired electoral boundaries. However these are applicable at the time of the Census (source year) and are not yet appropriate for the election (target year).

Denote year \(y\), with a Census falling on \(y_1\) and \(y_3\), and an election on year \(y_2\), and add this subscript to the Census variable estimate, \(\hat{C}_{t,y}\). To account for temporal changes, linear interpolation is used between Census years to get the final estimate of a Census variable for the target zone in the election year \(y_2\). This assumes that population evolves in a linear manner over time.

\[\hat{C}_{t,y_2} = \frac{y_3-y_2}{y_3-y_1}*\hat{C}_{t,y_1} + \frac{y_2-y_1}{y_3-y_1}*\hat{C}_{t,y_3}\]

\hypertarget{applied}{%
\subsubsection{Applied}\label{applied}}

Publically available Census data is aggregated and there are different resolutions accessible, ranging from SA1 (over 50,000 zones) to electoral divisions (150 zones). For this study electoral divisions are used as source zones, and this imputation method is applied for each of the 2004, 2007, 2010 and 2013 elections. To demonstrate its functionality, consider the imputation of socio-demographic variables for the electorate of Hume in New South Wales (NSW), at the time of the 2013 federal election. Figure \ref{fig:hume13} shows this region amongst other NSW electorates.

@@

{\centering \includegraphics[height=0.3\textheight]{skeleton_files/figure-latex/hume13-1}

@@

\caption{Some of the electoral boundaries in NSW for 2013, with the electoral boundary for Hume, shown in purple.}\label{fig:hume13}
\end{figure}

The Censuses neighbouring the 2013 election are those in 2011 and 2016, and the Hume boundary is changed, as seen by plotting the Hume boundary (purple) in the 2013 election over the divisions in 2016 (see Figure \ref{fig:hume16}).

@@

{\centering \includegraphics[height=0.3\textheight]{skeleton_files/figure-latex/hume16-1}

@@

There are many electorates in 2016 that intersect with the purple region (Hume boundary for 2013), these include the divisions of Riverina, Eden-Monaro and Hume, along with smaller intersecting areas with Fenner, Calare, Gilmore and Whitlam. To impute Census information for this purple region, calculate the percentage of each 2016 electorate that intersects with the purple region, which is then used to estimate intersection populations \(\hat{P}_{s,t}\).

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.17\columnwidth}\raggedright
Electorate (2016)\strut
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\raggedright
Percentage\strut
\end{minipage} & \begin{minipage}[b]{0.22\columnwidth}\raggedright
Population in Electorate\strut
\end{minipage} & \begin{minipage}[b]{0.40\columnwidth}\raggedright
Estimated Population Allocated to Purple Region: \(\hat{P}_{s,t}\)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.17\columnwidth}\raggedright
HUME\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
96.54\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
150643\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
145427\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.17\columnwidth}\raggedright
RIVERINA\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
25.11\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
155793\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
39117\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.17\columnwidth}\raggedright
EDEN-MONARO\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
11.09\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
147532\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
16358\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.17\columnwidth}\raggedright
CANBERRA\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.28\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
196037\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
548\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.17\columnwidth}\raggedright
FENNER\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.23\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
202955\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
474\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.17\columnwidth}\raggedright
WHITLAM\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.06\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
152280\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
92\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.17\columnwidth}\raggedright
GILMORE\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.06\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
150436\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
86\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.17\columnwidth}\raggedright
CALARE\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.01\%\strut
\end{minipage} & \begin{minipage}[t]{0.22\columnwidth}\raggedright
161298\strut
\end{minipage} & \begin{minipage}[t]{0.40\columnwidth}\raggedright
21\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

Now consider the socio-demographic \(AusCitizen\) - the proportion of people in the region who are Australian citizens.

>>>>

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Spatial modelling of the electoral two-party preferred vote in Australia. A study of federal elections between 2001 and 2016 via the eechidna R package.},
            pdfkeywords={federal election, Census, Australia, spatial modelling, imputation, socio-demographics, electorates, R, eechidna},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,]{biblatex}
\addbibresource{references.bib}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

@@

\begin{document}
\maketitle
\begin{abstract}
This research examines the relationships between electoral socio-demographic characteristics and two-party preference in Australian federal elections, with the six elections from 2001 to 2016 considered. Socio-demographic information is derived from the Australian Census, which occurs every five years. Since a Census is not directly available for each election, a method of spatio-temporal imputation is employed to impute Census data for the electorates in 2004-2013 elections. This accounts for both spatial and temporal changes in electoral characteristics between Censuses. To capture any spatial heterogeneity, a spatial error model is estimated for each election, which incorporates a spatially structured random effect vector that can thought of as the unobserved political climate in each electorate. Over time, most socio-demographic characteristics that affect electoral two-party preference do not vary, with industry of work, incomes, household mobility and de facto relationships having strong effects in the six elections. Education and unemployment are amongst those that have varying effects. It is also found that between 2004 and 2013, the spatial effect is insignificant, meaning that electorates voted effectively independently. All data featured in this study has been contributed to the \texttt{eechidna} \texttt{R} package (available on CRAN).
\end{abstract}
\begin{keywords}
federal election, Census, Australia, spatial modelling, imputation, socio-demographics, electorates, R, eechidna
\end{keywords}

@@

Australia has changed in many ways over the last two decades. Rising house prices, country-wide improvements in education, an ageing population, and a decline in religious affiliation, are just a few facets of the country's evolving socio-demographic characteristics. At the same time, political power has moved back and forth between the two major parties. In the 2007 and 2010 federal elections, the Australian Labor Party (Labor) was victorious, whereas the 2001, 2004, 2013 and 2016 elections were won by the Liberal National coalition (Liberal). The two-party preferred vote, a measure of support between these two parties, fluctuated between 47.3\% and 53.5\% (in favour of the Liberal party) over this period. This study explores how electoral socio-demographic characteristics relate to two-party preference, and whether their effects have changed over time. Electoral socio-demographics are derived from the Australian Census, and vote counts are obtained from Australian federal elections.

Joining these two data sources is problematic as there in an inherent asynchronicity in the two types of events. A Census is conducted by the Australian Bureau of Statistics (ABS) every five years, whereas a federal election (conducted by the Australian Electoral Commission (AEC)) usually occurs every three years. The first problem addressed is that of constructing appropriate Census data for the 2004, 2007, 2010 and 2013 elections - election years in which a Census does not occur. The predominant approach in previous studies is to join voting outcomes to the nearest Census, without accounting for any temporal differences (see \textcite{DavisStimson98}, \textcite{Stimson06}, \textcite{Liao09} and \textcite{Stimson09}). Furthermore, electoral boundaries change regularly, so spatial discrepancies also arise when matching with electoral data. To obtain appropriate Census data for these four elections, electoral socio-demographics are constructed using a spatio-temporal imputation that combines areal interpolation \autocite{Goodchild1993} and linear time-interpolation. Collecting and wrangling the raw data, along with the imputation process, are detailed in Section \ref{data}. All data and associated documentation relating to this procedure are available in the \texttt{eechidna} \texttt{R} package \autocite{eechidna}, providing a resource for future analysis .

Previous work on modelling Australian federal elections have found that aggregate socio-demographics are relatively good predictors of voting outcomes. \textcite{Forrest01} use multiple regression to model the Liberal and Labor primary vote for polling booths in the Farrer electorate in 1998 as a function of Census variables from 1996. \textcite{Stimson06}, \textcite{Stimson09} and \textcite{Stimson12} use principal component analysis of polling booths in the 2001, 2004 and 2007 elections respectively, also finding that socio-demographic characteristics of polling booths are linked to their two-party preferred vote. On the contrary, \textcite{Stimson09} models the polling booth swing vote (change in the two-party preferred vote) in the 2007 election, finding that little of swing vote can be explained by Census data. Instead of analyzing a single election in isolation, this paper employs a consistent model framework across six elections so that temporal changes in the effects of socio-demographics can be observed, where each federal elections is modelled with a cross-sectional dataset. The use of a regression framework to examine these socio-political relationships over time is seemingly absent from previous Australian studies. It also appears that no study has attempted any type of statistical analysis of socio-demographics in conjunction with voter behaviour in Australia since 2007, making this paper distinctly different from those previous.

The cross-sectional dataset for each election used here consists of the two-party preferred vote (response variable), and a set of common socio-demographic variables (explanatory variables) that characterize each electorate. To obtain these datasets, socio-demographic variables are first standardized, and then a principal component analysis is used to group variables into ``factors''. To account for the inherent spatial structure of the data, a spatial error model is then fit for each election. These steps are discussed in Section \ref{modelling}. In Section \ref{results} inference is conducted to see which effects are significant and how these effects change over time.

Section \ref{conclusion} concludes, followed by two supplementary sections.

@@

The voting outcome of interest is the electoral two-party preferred vote, which is provided by the Australian Electoral Commission (AEC) for the 2001, 2004, 2007, 2010, 2013 and 2016 elections via the AEC Tally Room. The AEC divides Australia into 150 regions, called electorates, with each corresponding to a single seat in the House of Representatives. Voting is compulsory in Australia, and each voter assigns a numbered preference to each available candidate in their electorate. The two-party preferred vote is determined by a tally of these preferences where, by convention, only the ranks of the Labor and Liberal candidates are considered. This is recorded as a percentage preference in favour of the Liberal party.

Socio-demographic variables are derived from the Census of Population and Housing (Census), which is a survey of every household in Australia, recording information such as age, gender, ethnicity, education level and income. There have been four Censuses so far in the 21st century, conducted in 2001, 2006, 2011 and 2016. The Australian Bureau of Statistics (ABS) conducts the Census and publishes aggregated information. The ABS uses electoral boundaries as defined by the AEC at the time of each Census, which may not match those in place at the subsequent election. From the available Census information aggregated at the electorate level, 65 socio-demographic variables are defined for each of the electorates to be used in the analysis.

Raw data is sourced online from the AEC and ABS websites in .csv and .xlsx files. The formats of these files differ over the years, making extracting the appropriate information a big task. The functions available in the \texttt{dplyr} \autocite{dplyr} and \texttt{readxl} \autocite{readxl} \texttt{R} packages are very useful, as they provide fast consistent tools for data manipulation and functions to import .xlsx files (respectively). The 2001 and 2006 Census data are published in a format where the information for each electorate is held in a separate document making it difficult to use the \texttt{dplyr} tools. Instead, cells have to be selected from each individual file to construct the desired variables. All scripts required for the data wrangling process can be found in the github repository for the \texttt{eechidna} \texttt{R} package \autocite{eechidna}, along with the raw data. The \texttt{eechidna} package makes this study entirely reproducible and provides a resource to help wrangle data for future Censuses and elections, when they become available.

@@

Between 2001 and 2016 there were six elections and four Censuses (see Figure \ref{fig:timeline}). Electoral boundaries are redistributed regularly by the AEC, meaning that only in the years where both a Census and election occur will the boundaries match - the case for the 2001 and 2016 elections. Therefore, for the four elections between 2004 and 2013, both temporal and spatial differences in electorates need to be accounted for when joining the electoral two-party preferred vote with Census data. For these elections a spatio-temporal imputation method is employed to obtain electoral socio-demographics. This method uses Census information from both before and after the election of interest.

@@

{\centering \includegraphics[width=0.9\linewidth]{draft-paper_files/figure-latex/timeline-1}

@@

For each election, neighbouring Census information has to be combined in some way so that it represents the boundaries in place at the time of the election. This is done by taking the electoral boundaries and imputing the corresponding socio-demographic characteristics for each of neighbouring Censuses, thereby addressing the spatial aspect. Next, to deal with the temporal component, characteristcs at the time of the election are constructed using linear interpolation between the spatially imputed neighbouring Census variables.

To account for spatial differences, the piece-wise approximation method in \textcite{Goodchild1993} is adopted. Consider a map of source zones \(s = 1,...,S\), for which socio-demographic information is available, and a set of target zones \(t = 1,...,T\) for which information is to be imputed. This is described in the context of a single election, and a single neighbouring Census.

Let the map of electoral boundaries at the time of a Census define the source zones, and let the boundaries at the time of the election be the target zones. Denote the area of intersection between a source zone \(s\) and a target zone \(t\) as \(A_{s,t}\). Additionally, let the population of the source zone \(s\) be \(U_s\) and the population of intersection between source zone \(s\) and target zone \(t\) be \(P_{s,t}\). The estimated population of intersection is given by:

\[\hat{P}_{s,t} = \frac{U_s*A_{s,t}}{\sum_{t=1}^T A_{s,t}}, \mbox{ for all } s=1,...,S \mbox{ and } t=1,...,T.\]
Note that this estimator implicitly assumes that populations are uniformly distributed within each source zone.

In order to calculate socio-demographic information for each of the target zones, a weighted average is taken using the estimated intersection populations as weights. Denote a given Census variable for the target zone by \(C_t\), and the same Census variable for the source zone as \(D_s\). Then, estimate \(C_t\) using:

\[\hat{C}_t = \frac{\sum_{s=1}^{S}{D_s*\hat{P}_{s,t}}}{\sum_{s=1}^{S}{\hat{P}_{s,t}}}, \mbox{ for each } t=1,...,T.\]
This concludes the spatial imputation of the socio-demographic characteristics for one target zone (a single electoral boundary), at time of only one of the neighbouring Censuses. This process is repeated for all of the target zones, and then for the other neighbouring Census.

To account for temporal changes, linear interpolation is used between Census years to get the final estimate of a Census variable for the target zone in the election year. Let \(y_1\) be the year of the Census preceeding an election, let \(y_2\) be the year of the election and \(y_3\) be the year of the Census that follows. Add this year subscript to the Census variable estimate \(\hat{C}_t\), resulting in \(\hat{C}_{t,y}\). Linear interpolating between these Census years results an imputed value for the election year, given by

\[\hat{C}_{t,y_2} = \frac{y_3-y_2}{y_3-y_1}*\hat{C}_{t,y_1} + \frac{y_2-y_1}{y_3-y_1}*\hat{C}_{t,y_3}.\]
Implicitly this assumes that population characteristics change in a linear manner over time.

\hypertarget{an-illustration-of-the-spatio-temporal-imputation}{%
\subsubsection{An illustration of the spatio-temporal imputation}\label{an-illustration-of-the-spatio-temporal-imputation}}

Census data is publicly available that different levels of aggregation, ranging from SA1 (over 50,000 zones) to electoral divisions (150 zones). For this study electoral divisions are used as source zones, and the imputation method is applied to produce socio-demographic variables for each of the 2004, 2007, 2010 and 2013 elections. As mentioned earlier, there is no need to impute socio-demographic variables for the 2001 and 2016 elections. To illustrate the method, consider the imputation of socio-demographic variables for the electorate of Hume in New South Wales (NSW) at the time of the 2013 federal election. The boundaries shown in Figure \ref{fig:hume13} define all target zones in NSW for 2013, with the target zone of interest (Hume) shaded purple.

@@

{\centering \includegraphics[height=0.3\textheight]{draft-paper_files/figure-latex/hume13-1}

@@

\caption{Some of the electoral boundaries in NSW for 2013, with the electoral boundary for Hume shown in purple.}\label{fig:hume13}
\end{figure}

The corresponding source zones from the 2016 Census are shown in Figure \ref{fig:hume16}. As can be seen, the Hume boundary from the 2013 election (shaded purple) does not perfectly match any of the source zones.

@@

{\centering \includegraphics[height=0.3\textheight]{draft-paper_files/figure-latex/hume16-1}

@@

There are many source zones from the 2016 Census that intersect with this purple region, including the divisions of Riverina, Eden-Monaro and Hume, along with smaller intersecting areas with Fenner, Calare, Gilmore and Whitlam. The proportion of each source zone that overlaps with the purple region is calculated, and used to obtain the intersecting populations \(\hat{P}_{s,t}\).

\footnotesize

\begin{longtable}[]{@{}llll@{}}
\toprule
\begin{minipage}[b]{0.21\columnwidth}\raggedright
Source Zone (2016)\strut
\end{minipage} & \begin{minipage}[b]{0.10\columnwidth}\raggedright
Proportion\strut
\end{minipage} & \begin{minipage}[b]{0.11\columnwidth}\raggedright
Source Zone Population\strut
\end{minipage} & \begin{minipage}[b]{0.46\columnwidth}\raggedright
Population Allocated to Purple Region: \(\hat{P}_{s,t}\)\strut
\end{minipage}\tabularnewline
\midrule
\endhead
\begin{minipage}[t]{0.21\columnwidth}\raggedright
HUME\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.9654\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
150643\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
145427\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.21\columnwidth}\raggedright
RIVERINA\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.2511\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
155793\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
39117\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.21\columnwidth}\raggedright
EDEN-MONARO\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.1109\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
147532\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
16358\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.21\columnwidth}\raggedright
CANBERRA\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.028\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
196037\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
548\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.21\columnwidth}\raggedright
FENNER\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.0023\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
202955\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
474\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.21\columnwidth}\raggedright
WHITLAM\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.0006\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
152280\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
92\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.21\columnwidth}\raggedright
GILMORE\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.0006\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
150436\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
86\strut
\end{minipage}\tabularnewline
\begin{minipage}[t]{0.21\columnwidth}\raggedright
CALARE\strut
\end{minipage} & \begin{minipage}[t]{0.10\columnwidth}\raggedright
0.0001\strut
\end{minipage} & \begin{minipage}[t]{0.11\columnwidth}\raggedright
161298\strut
\end{minipage} & \begin{minipage}[t]{0.46\columnwidth}\raggedright
21\strut
\end{minipage}\tabularnewline
\bottomrule
\end{longtable}

\normalsize

Now consider the socio-demographic variable \(AusCitizen\) - the proportion of people in the region who are Australian citizens.

\footnotesize
